# PTR-abundance correlations

We start by reading in the PTR data and merging it with the OD data and the read counts.

```{r}
library(data.table)
library(ggplot2)
theme_minimal() |> theme_set()

ptrs <- fread("data/rates.csv") |> melt(variable.name="sample_id", value.name="log2PTR")
names(ptrs)[1] <- "reference"
ptrs[, "well" := tstrsplit(sample_id, "_")[[1]]]
ods <- fread("data/OD_endpoints.csv")
abundances <- fread("data/reference_counts.csv")
merged <- ods[abundances[ptrs, on="sample_id"], on="well"][!is.na(log2PTR)]
merged[, "log_ratio" := 0.5*log(NC_000913.3+1) - 0.5 * log(phiX174+1)]
cuts <- c(225, 300, 510)
merged[, "group" := "stationary"]
merged[minutes <= cuts[1], "group" := "acceleration"]
merged[minutes > cuts[1] & minutes <= cuts[2], "group" := "mid-log"]
merged[minutes > cuts[2] & minutes <= cuts[3], "group" := "deceleration"]
merged[, "group" := factor(group, levels=c("acceleration", "mid-log", "deceleration", "stationary"))]
merged
```
Let's have a look at the PTRs over time.

```{r}
ggplot(merged) +
  aes(x=minutes/60, y=log2PTR, color=factor(offset), group=1) +
  stat_smooth(span=1/2) +
  geom_point() +
  labs(x="time [h]", y="replication rate [log₂ PTR]", color="offset [min]")
ggsave("figures/ptr_over_time.png", dpi=300)
```

Let's look at E. coli replication rates against E. coli abundance inferred from sequencing alone.

```{r}
ggplot(merged) +
  aes(x=log_ratio, y=log2PTR, group=1) +
  stat_smooth() +
  geom_point() +
  labs(x="E. coli abundance [centered log-ratio]", y="replication rate [log₂ PTR]")
  geom_point()
ggsave("figures/ptr_vs_clr.png", dpi=300)
```
And separated by group.

```{r, fig.width=8, fig.height=3}
ggplot(merged) +
  aes(x=log_ratio, y=log2PTR, color=group) +
  stat_smooth(method="lm") +
  geom_point() +
  labs(x="E. coli abundance [centered log-ratio]", y="replication rate [log₂ PTR]") +
  facet_wrap(~ group, nrow=1, scale="free_x") +
  guides(color = FALSE) +
  theme(panel.spacing.x = unit(6, "mm"))
ggsave("figures/ptr_vs_clr_groups.png", dpi=300, width=8, height=3)
```

With the corresponding tests.

```{r, fig.width=4, fig.height=4}
pearson = function(x, y) {
  res <- cor.test(x, y)
  return(list(
    r = res$estimate,
    n = res$parameter,
    p = res$p.value,
    conf_low = res$conf.int[1],
    conf_high = res$conf.int[2]
  ))
}

corrs <- merged[, pearson(log_ratio, log2PTR), by="group"]

ggplot(corrs) +
  aes(x=group, y=r, ymin=conf_low, ymax=conf_high, color=group) +
  geom_hline(yintercept = 0, linetype="dashed", color="black") +
  geom_linerange() +
  geom_point(size=2) +
  labs(x="", y="Pearson r") +
  guides(color = FALSE) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

ggsave("figures/ptr_vs_clr_corrs.png", dpi=300, width=4, height=4)
print(corrs)
```

And we can do the same for the OD data.

```{r}
ggplot(merged) +
  aes(x=log2(OD600), y=log2PTR) +
  stat_smooth(span=1/3) +
  geom_point() +
  labs(x="log₂ OD600", y="replication rate [log₂ PTR]")
  geom_point()
ggsave("figures/ptr_vs_od.png", dpi=300)
```

And separated by group.

```{r}
ggplot(merged) +
  aes(x=log2(OD600), y=log2PTR, color=group) +
  stat_smooth(method="lm") +
  geom_point() +
  labs(x="log₂ OD600", y="replication rate [log₂ PTR]") +
  facet_wrap(~ group, scale="free_x", nrow=1)
ggsave("figures/ptr_vs_od_groups.png", dpi=300)
```

And the tests.

```{r}
merged[, pearson(log2(OD600), log2PTR), by="group"]
```